Learning a functional distance between organizations with unsupervised embeddings of scientific mobility

### 2020-07-14 paper/current/main.tex
update first section

How far apart are two places?
The question is surprisingly hard to answer when it involves human mobility.
While geographic distance may have driven human movements in past eras, it holds less importance to a world connected by rapid transit, complex international relations, and global trade.
For instance, a person currently living in Australia is more likely to migrate to the United Kingdom, a country with similar language and culture, than to closer and more populous countries like Indonesia. 
Similarly, a prospective student living in South Korea is more likely to attend university in Canada rather than the closer though diplomatically isolated North Korea. 
Despite its complexity, ubiquitous models of human mobility such as the Gravity~\autocite{zipf1946gravity} and Radiation~\autocite{simini2012universal} model continue to predict mobility as a function of geographic distance.
Although alternatives exist to quantify \textit{functional} distances based on factors other than geography~\autocite{boschma2005proximity, brown1970functional, brown1970migration, kim2018functional}, geographic distance remains the unambiguous and foundational quantity for modeling mobility.

Here, we propose a neural embedding model that uses real-world mobility trajectories to encode the complex landscape of global mobility into a dense, continuous, and meaningful vector-space representation, from which a \textit{functional distance} between places can be derived.
This approach, originally designed for learning language models~\autocite{mikolov2013word2vec} has been used to reveal insights in text~\autocite{linzhuo2020hyperbolic, tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2016twitch} and networks~\autocite{perozzi2014deepwalk, grover2016node2vec}.
In addition to providing a data-driven functional distance, this approach is computationally efficient, scalable for large datasets, robust to noise, and is readily interpretable through the application of simple vector arithmetic.

We use scientific mobility (scientific co-affiliation) as a case study to demonstrate the neural embedding approach. 
Scientific mobility is a central aspect of the increasingly-globalized scientific system~\autocite{czaika2018globalisation}, being strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation, armano2017innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale, jonkers2013return, franzoni2014advantage}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, azoulay2011diffusion, morgan2018prestige}.
For nations and organizations hoping to win in the global competition for talent~\autocite{box2008competition}, understanding the landscape of global scientific mobility is essential. 
However, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, vannoorden2012mobility, deville2014career, ackers2008mobility, scott2015dynamics}.
Neural embedding offers a means of capturing the complex global structure of scientific mobility in a dense and coherent representation that supports modeling and predicting mobility flows, mapping patterns of mobility at multiple scales, and quantifying abstract notions such as academic prestige.

The most fine-grained geographical unit of scientific production is the organization, such as an university, research institute, business, or government agency.
As a scientist's careers progress, they move between organizations or pick up additional simultaneous affiliations forming co-affiliation or \textit{mobility trajectories}.
These mobility trajectories, represented as ordered lists of organizations, can be derived from affiliations listed on the researcher's publication histories~\autocite{robinson2019mobility, sugimoto2017mostimpact} (Fig.~\ref{fig:methods:mobility_sentence}).
We use the trajectories of the three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.
A vector-space embedding of organizations is learned by using mobility trajectories as input to the standard skip-gram with negative sampling, or \textit{word2vec} neural-network architecture.
In the resulting embedding, each organization corresponds to a single vector, and vectors relate to one another based on the likelihood of organizations appearing near together in the mobility trajectory data. 
The functional distance between two organizations is calculated as the cosine distance between their vectors, termed the \textit{embedding distance}.

@@

Where the \textit{expected flux}, $\hat{T_{ij}}$, or expected number of authors co-affiliated with a pair organizations during the time period is proportional to the organization's sizes (the sum of both mobile and non-mobile authors, see Data and Methods), $m_{i}$ and $m_{j}$, and decays as a function of their distance, $f(r_{ij})$.
The distance function is defined in terms of either the geographic distance between organizations or the embedding distance. 
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).
We argue that the performance of a distance measure can be observed from its ability to explain expected mobility between organizations and to make predictions using the gravity model.

>>>>

How far apart are two places?
The question is surprisingly hard to answer when it involves human mobility.
While geographic distance may have driven human movements in past eras, it holds less importance to a world connected by rapid transit, complex international relations, and dense airline networks.
For instance, a person currently living in Australia is more likely to migrate to the United Kingdom, a country with similar language and culture, than to closer and more populous countries like Indonesia. 
Similarly, a prospective student living in South Korea is more likely to attend university in Canada rather than the closer though diplomatically isolated North Korea. 
Despite its complexity, geographic distnace is used as the basis for the most prominent models of mobility such as the Gravity~\autocite{zipf1946gravity} and Radiation~\autocite{simini2012universal}.
So-called \textit{functional distances}~\autocite{boschma2005proximity, brown1970functional, brown1970migration, kim2018functional} have been proposed that attempt to learn alternative notions of distance from real-world data or \textit{a priori} relationships between geographic entities.

Here, we propose a neural embedding model that uses real-world mobility trajectories to encode the complex landscape of global mobility into a dense, continuous, and meaningful vector-space representation, from which a \textit{functional distance} between places can be derived.
This approach, originally designed for learning language models~\autocite{mikolov2013word2vec} has been used to reveal insights in text~\autocite{linzhuo2020hyperbolic, tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2016twitch} and networks~\autocite{perozzi2014deepwalk, grover2016node2vec}.
Neural embedding is also computationally efficient, scalable for large datasets, robust to noise, and is readily interpretable through the application of simple vector arithmetic.

We use scientific mobility, inferred from affiliations, to demonstrate the neural embedding approach. 
Scientific mobility is a central aspect of the increasingly-globalized scientific system~\autocite{czaika2018globalisation, altbach2004globalization}, being strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation, armano2017innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, azoulay2011diffusion, morgan2018prestige}.
For nations and organizations hoping to win in the global competition for talent~\autocite{box2008competition}, understanding the landscape of global scientific mobility is essential. 
However, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, vannoorden2012mobility, deville2014career, ackers2008mobility, scott2015dynamics}.
One roadblock is the lack of a meaningful measure of functional distance between countries, cities, and organizations.

Here, scientific mobility is operationalized using researcher's organizational affiliations derived from their publication histories indexed in a large-scale bibliographic database. 
The most fine-grained geographical unit in this data is the organization, such as an university, research institute, business, or government agency.
As a scientist's careers progress, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:methods:mobility_sentence}.
These trajectories encodie both asynchronous migration, and synchronous mobility---the holding of multiple simultaneous co-affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, markova2016synchronous, sugimoto2017mostimpact} (see Supplementary Materials).
We use the trajectories of the three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.

A vector-space embedding of organizations is learned by using mobility trajectories as input to the standard skip-gram with negative sampling, or \textit{word2vec} neural-network architecture.
In the resulting embedding, each organization corresponds to a single vector, and vectors relate to one another based on the likelihood of organizations appearing near together in the mobility trajectory data. 
The functional distance between two organizations is calculated as the cosine distance between their vectors, termed the \textit{embedding distance}.

@@

Where the \textit{expected flux}, $\hat{T_{ij}}$, or expected number of authors who were affiliated with both organizations during the time period is proportional to the organization's sizes (the sum of both mobile and non-mobile authors, see Data and Methods), $m_{i}$ and $m_{j}$, and decays as a function of their distance, $f(r_{ij})$.
The distance function is defined in terms of either the geographic distance between organizations or the embedding distance. 
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).
We argue that the performance of a distance measure can be observed from its ability to explain expected mobility between organizations and to make predictions using the gravity model.
